---
title: "Homework 3"
author: "Aji John"
date: "Winter Quarter 2018"
output:
  pdf_document: default
  word_document: default
---



```{r setup, include=FALSE}
### Setting up the packages, options we'll need:
library(knitr)
library(ggplot2)
library(tidyverse)
library(uwIntroStats)
knitr::opts_chunk$set(echo = FALSE)

```


# Question 1
Perform a statistical analysis evaluating an association between serum creatinine level and 5 year all-cause mortality by comparing geometric mean serum creatinine levels between groups defined by vital status at 5 years.   In your analysis, allow for heteroscedasticity.  Provide full statistical inference about an association between creatinine and 5 year all-cause mortality.   

```{r Q1a,echo=FALSE,include=FALSE}
### Question 1:
# Part (a)
mri <- read.table("data/mri.txt",header=TRUE) 
mri$deathin5 <- ifelse(mri$obstime <= 365.25 * 5 & mri$death == 1, TRUE, FALSE) 
mri$death5 <- (mri$obstime < 5*365) & (mri$death == 1)
mri$logcrt <- log(mri$crt)
#kable(descrip(mri[mri$death==0,'crt']), format = "markdown")
#kable(summary(cars), format = "markdown")
kable(mri%>%filter(!is.na(crt) & deathin5==FALSE) %>% 
        group_by(deathin5) %>%
        summarise(n=n(),sd=sd(crt),mean=mean(crt))%>%
        select(c('n','mean','sd')) %>%
        as.data.frame() , format = "markdown")
 
t.test(mri[!is.na(mri$crt)  & (mri$deathin5 ==FALSE),]$logcrt,mri[!is.na(mri$crt) &  (mri$deathin5 ==TRUE),]$logcrt,var.equal = FALSE)

#	Welch Two Sample t-test

#data:  mri[!is.na(mri$crt) & (mri$deathin5 == FALSE), ]$logcrt and mri[!is.na(mri$crt) & (mri$deathin5 == TRUE), ]$logcrt
#t = -4.245, df = 145.73, p-value = 3.874e-05
#alternative hypothesis: true difference in means is not equal to 0
#95 percent confidence interval:
# -0.19376718 -0.07065662
#sample estimates:
#  mean of x   mean of y 
#0.005763226 0.137975123 

```

Model we got is E[log(crt)|deathin5]=0.005763+0.132212×deathin5.

From linear regression analysis on log transformed creatinine level(CRT) using Huber-White estimates of the standard error, we estimate that between two groups defined by vital status at 5 years, the geometric mean CRT is 14.13% higher in the group which survives in 5 years. A 95% CI suggests that this observation is not unusual(?) if the true relationship between geometric means were such that the group which survives in 5 years geometric mean CRT were between 7.38% and 21.31% higher from the group that dies in 5 years. Because the two-sided P value is P < .0005, we reject the null hypothesis that there is no linear trend in the geometric mean CRT across the two groups.

F
* Two samples skipped as 'NA' featured in them.
* Units for mean is mg/dl

# Question 2
Perform a regression analysis evaluating an association between serum creatinine level and age by comparing mean serum creatinine levels across groups defined by age as a continuous variable.   In your analysis, allow for heteroscedasticity.  (Provide formal inference where asked to.) 
 
## (a) 
Provide a brief description of the statistical methods for the model you fit to address the question of an association between creatinine and age.

```{r Q1a}
### Question 1:
# Part (a)
plot(data=mri,crt~age)
abline(lm(crt~age, data = mri),col="red")

```
We first do the some exploratory plots to so some look into relationship. From the above figure it looks like there is weak linear relationship. Lets look closely by plotting the residuals. 

```{r Q1a}
### Question 1:
# Part (a)
agefit <- lm(formula = crt ~ age, data = mri)
plot(x=fitted(agefit),y=resid(agefit),ylab = "Residuals",xlab = "Fitted") 
abline(h=0,col="red")

```

By looking at the residulas, it confirms the same. Lets now look at the linear analysis output.

```{r Q1a,echo=FALSE,include=FALSE}
### Question 1:
# Part (a)
regress("mean",formula = crt ~ age, data = mri)

#Call:
#regress(fnctl = "mean", formula = crt ~ age, data = mri)

#Residuals:
#     Min       1Q   Median       3Q      Max 
#-0.61643 -0.19415 -0.04401  0.15599  2.86129 

#Coefficients:
#                 Estimate   Naive SE   Robust SE     95%L       95%H         F stat    df Pr(>F)   
#[1] Intercept      0.6485     0.1531     0.1897        0.2760     1.021          11.68 1    0.0007 
#[2] age           5.571e-03  2.047e-03  2.586e-03     4.941e-04  0.01065          4.64 1    0.0315 

#Residual standard error: 0.3016 on 731 degrees of freedom
#  (2 observations deleted due to missingness)
#Multiple R-squared:  0.01003,	Adjusted R-squared:  0.008674 
#F-statistic: 4.641 on 1 and 731 DF,  p-value: 0.03154

#Call:
#lm(formula = crt ~ age, data = mri)

#Residuals:
#     Min       1Q   Median       3Q      Max 
#-0.61643 -0.19415 -0.04401  0.15599  2.86129 

#Coefficients:
#            Estimate Std. Error t value Pr(>|t|)    
#(Intercept) 0.648490   0.153094   4.236 2.57e-05 ***
#age         0.005571   0.002047   2.721  0.00666 ** 
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#Residual standard error: 0.3016 on 731 degrees of freedom
#  (2 observations deleted due to missingness)
#Multiple R-squared:  0.01003,	Adjusted R-squared:  0.008674 
#F-statistic: 7.405 on 1 and 731 DF,  p-value: 0.00666

```

“From linear regression analysis using Huber-White estimates of the standard error, we estimate that for each year difference in age between two populations, the difference in mean DSST is 0.863 points lower in the older population. A 95% CI suggests that this observation is not unusual if the true difference in mean DSST were between .715 and 1.01 points lower per year difference in age. Because the two sided P value is P < .0005, we reject the null hypothesis that there is no linear trend in the average DSST across age groups.”

* Two samples skipped as 'NA' featured in them.
* Units for mean is mg/dl

## (b)
What are the sample size, sample mean and sample standard deviation of creatinine levels among subjects who died within 5 years? 
```{r Q1b, fig.align='center'}
# Part (b)

kable(mri%>%filter(!is.na(crt) & deathin5==TRUE) %>% 
        group_by(deathin5) %>%
        summarise(n=n(),sd=sd(crt),mean=mean(crt))%>%
        select(c('n','mean','sd')) %>%
        as.data.frame() , format = "markdown")

```
* Units for mean is mg/dl

## (c)
What are the point estimate, the estimated standard error of that point estimate, and the 95% confidence interval for the mean creatinine level in a population of similar subjects who would survive at least 5 years?

```{r , include=FALSE,fig.align='center'}
# Part (c)

t.test(mri[mri$deathin5==FALSE,]$crt)

crt_0 <- mri[!is.na(mri$crt) & mri$deathin5==FALSE,]$crt

xbar<- mean(crt_0)
sigma <- sd(crt_0)

e1 <- qnorm(.975) * (sigma /sqrt(length(crt_0))) # error
e2 <- qnorm(.025) * (sigma /sqrt(length(crt_0)))

SE <- (sigma /sqrt(length(crt_0)))

int.s1 <- xbar - e1
int.s2 <- xbar - e2

```

Point estimate is 1.033 mg/dl
SE is 0.009974404
95% CI is 1.01 to 1.05

## (d)
What are the point estimate, the estimated standard error of that point estimate, and the 95% confidence interval for the mean creatinine level in a population of similar subjects who would die within 5 years? 

```{r ,include=FALSE, fig.align='center'}
# Part (d)

t.test(mri[mri$deathin5==TRUE,]$crt)

crt_1 <- mri[!is.na(mri$crt) & mri$deathin5==TRUE,]$crt

xbar<- mean(crt_1)
sigma <- sd(crt_1)

e1 <- qnorm(.975) * (sigma /sqrt(length(crt_1))) # error
e2 <- qnorm(.025) * (sigma /sqrt(length(crt_1)))

SE <- (sigma /sqrt(length(crt_0)))

int.s1 <- xbar - e1
int.s2 <- xbar - e2

```
Point estimate is 1.2157 mg/dl
SE is 0.01918
95% CI is 1.13 to 1.30

## (e)
What are the point estimate and the 95% confidence interval for the difference in creatinine means between a population of similar subjects that survives at least 5 years and a population of similar subjects that dies within 5 years? What is the P value for testing the hypothesis that the two populations have the same mean creatinine level? What conclusions do you reach about a statistically significant association between serum creatinine and 5 year all-cause mortality?

```{r , include=FALSE,fig.align='center'}
# Part (e)

t.test(mri[!is.na(mri$crt)  & (mri$deathin5 ==FALSE),]$crt,mri[!is.na(mri$crt) &  (mri$deathin5 ==TRUE),]$crt,var.equal = TRUE)

#	Two Sample t-test

#data:  mri[!is.na(mri$crt) & (mri$deathin5 == FALSE), ]$crt and mri[!is.na(mri$crt) & (mri$deathin5 == TRUE), ]$crt
#t = -6.1798, df = 731, p-value = 1.066e-09
#alternative hypothesis: true difference in means is not equal to 0
#95 percent confidence interval:
# -0.2394432 -0.1239879
#sample estimates:
#mean of x mean of y 
# 1.033987  1.215702 
```
The best estimate of the difference, i.e. the point estimate is -0.18175 units ( Survive Sample, Observed to Die Sample.  )

95% CI , difference of mean 'crt' between the groups is (-0.2394432 , -0.1239879).

p-value is 1.066e-09, and is found to be significant, and we reject our null hypothesis that the two populations have the same mean creatinine levels.


## (f)
Although we did not consider age at the time of enrollment in the questions above, could the association analysis for creatinine level and 5 year all-cause mortality conducted using the t-test potentially be confounded by the age of the subjects at the time of the MRI?  Briefly explain why or why not this is plausible?  Provide any descriptive statistics (e.g. an appropriate table, plot, etc.) giving evidence for or against the association results and conclusions above with the t-test potentially being confounded by age.

```{r ,fig.align='center'}
# Part (e)

boxplot(age~deathin5,mri, main="Age distribution", 
  	xlab="5 Year all-cause mortality", ylab="Age (Yrs)",col=topo.colors(2))

legend("topright", inset=.02, title="Groups",
   c("TRUE-Survived","FALSE"), fill=topo.colors(2), horiz=TRUE, cex=0.8)


```

The age distribution is notable in the study, we can see from the above figure that median is not the same for the two groups, and it is possibly adding the variation. 


# Question 2

Perform statistical analyses evaluating an association between serum creatinine and 5 year all-cause mortality by comparing mean creatinine levels across groups defined by vital status at 5 years using linear regression that presumes homoscedasticity. As this problem is directed toward illustrating correspondences between the t test and linear regression, you do not need to provide full statistical inference for this problem. Instead, just answer the following questions.

## (a)

Fit a regression model where the response variable is creatinine level and the predictor variable is an indicator variable for a subject dying within 5 years (i.e., a value of 1 if subject died within 5 years, and a value of 0 if the subject survived at least 5 years).   Provide an interpretation of the intercept and slope of this regression model.

```{r ,include=FALSE}
### Question 2:
# note: round() Does not print trailing zeros
lm(data=mri,crt~deathin5)
```

Our model came out to be 'crt = 0.18 * deathin5True + 1.03

So, when the deathin5True is 0, i.e. alive in 5 years, the mean 'crt' level is 1.03 mg/dl.  For two groups which differ by one unit ( here only two groups - Dead vs Alive in 5 years).

Estimated trend in mean crt by subject dying or alive within 5 years is an average difference of 0.1817 mg/dl between two groups (crt is higher in individuals who died within 5 years)

```{r}
### Question 2:
# note: round() Does not print trailing zeros
plot(data=mri,crt~deathin5)
abline(lm(crt~deathin5, data = mri))
```
## (b)

Is the regression model you fit a saturated model? Briefly explain why or why not. 

```{r,include=FALSE}
### Question 2:
# note: round() Does not print trailing zeros
plot(data=mri,mri$crt)
```

The regression model is saturated as E[Yi|X1i = 0], and E[Yi|X1i = 1]. The deathin5 is a binary variable and can have only (0,1).

## (c)

Using the parameter estimates from the regression model, what is the estimate of the mean creatinine level in a population of similar subjects who would die within 5 years? How does this compare to the corresponding estimate from problem 1? If there are any differences, explain the source of the differences.

```{r,include=FALSE}
### Question 2:
# note: round() Does not print trailing zeros
lm(data=mri,crt~deathin5)
```
Our model came out to be - 'crt = 0.18 * deathin5True + 1.03

So, Mean creatinine level for individuals whol would die in 5 years comes to be 0.18 * 1 + 1.03 = 1.21 mg/dl. It is same as what was found in problem 1.

## (d)

Using the parameter estimates from the regression model, what is an estimate of the mean creatinine level in a population of similar subjects who would survive at least 5 years? How does this compare to the corresponding estimate from problem 1? If there are any differences, explain the source of the differences.

Our model came out to be - 'crt = 0.18 * deathin5True + 1.03

So, Mean creatinine level for individuals who would survive in 5 years comes to be 0.18 * 0 + 1.03 = 1.03 mg/dl. It is same as what was found in problem 1.

## (e)
Using the results from the regression model, provide a 95% confidence interval for mean creatinine level among a population of subjects who survive at least 5 years? How does this compare to the corresponding estimate from problem 1? If there are any differences, explain the source of the differences.

Calculating the CI

```{r,include=TRUE}
### Question 2:
# note: round() Does not print trailing zeros
kable(confint(lm(crt~deathin5==0, data = mri)), format = "markdown")
```
There is difference 

## (f)
If we presume the variances are equal in the two populations, what is the regression based estimates of the standard deviation and the variance within each group from the regression model? 

SD within each group is 0.296, variance is 0.087616.


## (g)
Using the parameter estimates from the regression model, what are the point estimate, the estimated standard error of the point estimate, and the 95% confidence interval for the difference in means between a population that survives at least 5 years and a population that dies within 5 years? What is the P value for testing the hypothesis that the two populations have the same mean creatinine level? What conclusions do you reach about a statistically significant association between serum creatinine level and 5 year all-cause mortality? How does this compare to the corresponding inference from problem 1?

Parameter estimates are 1.03 and 0.18 , and we estimate that between the groups, the mean difference in crt is 0.18, higher for the group that dies within 5 years.

SE of the point estimate -  0.2956

95% CI for the difference in means for the studied groups is  0.12 to 0.23.

P-value is 1.066e-09 which is < 0.0005, we reject the null hypothesis that there is no difference in creatinine leves between the studied groups.
 
# Question 3
 
Perform statistical analyses evaluating an association between serum creatinine level and 5 year all-cause mortality by comparing mean creatinine levels across groups defined by vital status at 5 years using a t-test that allows for heteroscedasticity (i.e., the possibility of unequal variances across groups).  What is the P value for testing the hypothesis that the two populations have the same mean creatinine level? What conclusions do you reach about a statistically significant association between serum creatinine and 5 year all-cause mortality? How do the results of this analysis differ from the results from problem 1 where a t-test that presumes homoscedasticity is used? If there are any differences, explain the source of the differences.  (Again, we do not need a formal report of the inference.)

```{r , include=FALSE,fig.align='center'}
# Part (e)

t.test(mri[!is.na(mri$crt)  & (mri$deathin5 ==FALSE),]$crt,mri[!is.na(mri$crt) &  (mri$deathin5 ==TRUE),]$crt,var.equal = FALSE)

#	Welch Two Sample t-test

#data:  mri[!is.na(mri$crt) & (mri$deathin5 == FALSE), ]$crt and mri[!is.na(mri$crt) & (mri$deathin5 == TRUE), ]$crt
#t = -4.1052, df = 132.97, p-value = 7.011e-05
#alternative hypothesis: true difference in means is not equal to 0
#95 percent confidence interval:
# -0.26927031 -0.09416079
#sample estimates:
#mean of x mean of y 
# 1.033987  1.215702 

```
In case of heteroscedasticity t-test, P-value = 7.011e-05 which is different from problem 1 where it is p-value = 1.066e-09. Default assumption in t-test is unequal variance. The conclusion of the test is same as the results are still statistically significant. The differences are in CIs and P-value. The differences in CI because we used pooled SD. P-value changes because of the calculated t-statistic is different.
 
# Question 4

Perform statistical analyses evaluating an association between serum creatinine and 5 year all-cause mortality by comparing mean creatinine levels across groups defined by vital status at 5 years using a linear regression model that allows for heteroscedasticity. What is the P value for testing the hypothesis that the two populations have the same mean creatinine level? What conclusions do you reach about a statistically significant association between serum creatinine and 5 year all-cause mortality? How do the results of this analysis differ from the results in problem 3 where a t-test that allows for heteroscedasticity is used? If there are any differences, explain the source of the differences. (Again, we do not need a formal report of the inference.) 


```{r ,include=FALSE}
### Question 2:
# note: round() Does not print trailing zeros
robustm1 <- regress("mean",crt~deathin5,data=mri)
robustm1
```

We use robust error which allows for heteroscedasticity, and p-value: 4.296e-05. We still infer the same conclusion that the difference in mean crt between the groups exists, and is found to be significant.  it is still significant. P-value is different, but  CI is for difference in mean is same (0.09504    0.2684) vs (0.09416079,0.26927031)