---
title: "Homework 4"
author: "Aji John"
date: "Winter Quarter 2018"
output:
  pdf_document: default
  word_document: default
---



```{r setup, include=FALSE}
### Setting up the packages, options we'll need:
library(knitr)
library(ggplot2)
library(tidyverse)
library(uwIntroStats)
knitr::opts_chunk$set(echo = FALSE)

```

# Retrospective highcrt ~ deathin5

   ## odds for highcrt for those who are alive in 5 
   
# Prospective deathin5 ~ highcrt
    ## odds for those who are dead in 5 /vs alive
    
# Question 1
Perform a regression analysis for evaluating an association between serum creatinine level and 5 year all-cause mortality by comparing the odds of death within 5 years across groups defined by whether the subjects have “high”  or “low” creatinine levels, where serum creatinine levels greater than 1.2 are considered to be “high”, (i.e., “high” corresponds to  creatinine > 1.2 mg/dl and “low” corresponds to creatinine less than or equal 1.2 mg/dl ).  In your logistic regression model, use an indicator of death within 5 years as the  response, and an indicator of high serum creatinine level as the predictor. (Only provide a formal report of inference when asked to.) 

```{r ,echo=FALSE,include=FALSE}
### Question 1:
# Part (a)
mri <- read.table("data/mri.txt",header=TRUE) 
mri$deathin5 <- ifelse(mri$obstime <= 365.25 * 5 & mri$death == 1, TRUE, FALSE) 
mri$death5 <- (mri$obstime < 5*365) & (mri$death == 1)

#create high/low for crt
# 1 - high , 0 - low
mri$crtind <- ifelse(mri$crt>1.2,1,0)
mri$logcrt <- log(mri$crt)
#kable(descrip(mri[mri$death==0,'crt']), format = "markdown")
#kable(summary(cars), format = "markdown")
kable(mri%>%filter(!is.na(crt) & deathin5==FALSE) %>% 
        group_by(deathin5) %>%
        summarise(n=n(),sd=sd(crt),mean=mean(crt))%>%
        select(c('n','mean','sd')) %>%
        as.data.frame() , format = "markdown")
 

```

## (a)

Is this a saturated regression model? Explain your answer.

Yes, it is a saturated model, as number of parameters equal to number of coefficients.

## (b)

Provide an interpretation of the slope and the intercept in your regression model  in terms of the response variable (indicator of death within 5 years) and the predictor variable (high  creatinine).

Odds when predictor is 0 – Found by exponentiation of the intercept from the logistic regression: exp(b0)

Odds ratio between groups differing in the value of the predictor by 1 unit – Found by exponentiation of the slope from the logistic regression: exp(b1)

## (c)
From the logistic regression model, what is the estimated odds of dying within 5 years for subjects with low creatinine levels.   What is the estimated probability of dying within 5 years from the logistic regression model for subjects with low creatinine levels ? 

```{r,echo=FALSE,include=FALSE}
### Question 1:
# Part (a)
logiscrtl <- glm(deathin5 ~ crtind,family = binomial,data = mri)
summary(logiscrtl)

#Call:
#glm(formula = deathin5 ~ crtind, family = binomial, data = mri)

#Deviance Residuals: 
#    Min       1Q   Median       3Q      Max  
#-0.7933  -0.5387  -0.5387  -0.5387   2.0009  

#Coefficients:
#            Estimate Std. Error z value Pr(>|z|)    
#(Intercept)  -1.8567     0.1225 -15.152  < 2e-16 ***
#crtind        0.8618     0.2148   4.012 6.03e-05 ***
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#(Dispersion parameter for binomial family taken to be 1)

#    Null deviance: 656.75  on 732  degrees of freedom
#Residual deviance: 641.50  on 731  degrees of freedom
#  (2 observations deleted due to missingness)
#AIC: 645.5

#Number of Fisher Scoring iterations: 4

# exp(--1.8567 )
#[1] 6.402573
#> 6.402573/(1+6.402573)
#[1] 0.8649118

#regress(fnctl = "odds", formula = deathin5 ~ crtind, data = mri)


#Call:
#regress(fnctl = "odds", formula = deathin5 ~ crtind, data = mri)

#Deviance Residuals: 
#    Min       1Q   Median       3Q      Max  
#-0.7933  -0.5387  -0.5387  -0.5387   2.0009  

#Coefficients:

#Raw Model:
#                 Estimate  Naive SE  Robust SE       F stat    df Pr(>F)   
#[1] Intercept     -1.857     0.1225    0.1227           228.96 1  < 0.00005
#[2] crtind         0.8618    0.2148    0.2151            16.05 1     1e-04 

#Transformed Model:
#                 e(Est)    e(95%L)   e(95%H)         F stat    df Pr(>F)   
#[1] Intercept      0.1562    0.1228    0.1987           228.96 1  < 0.00005
#[2] crtind         2.367     1.552     3.611             16.05 1     1e-04 

#(Dispersion parameter for binomial family taken to be 1)

#    Null deviance: 656.75  on 732  degrees of freedom
#Residual deviance: 641.50  on 731  degrees of freedom
#  (2 observations deleted due to missingness)
#AIC: 645.5

#Number of Fisher Scoring iterations: 4
 
```

Odds of dying within 5 years for subjects with low creatinine levels is e-1.8567=0.1562

Probability of dying within 5 years for subjects with low creatinine levels is 0.1350977.

 - Using  prob= odds / (1+odds):  0.1562 / (1+0.1562)= 0.1350977.

## (d)
For the subset of subjects in the sample with low serum creatinine, calculate the proportion who died within 5 years. Also calculate the sample odds of dying within 5 years for subjects with low creatinine levels.  Compare the sample proportion and sample odds  to the corresponding estimates in part 1c from the logistic regression model? Briefly explain any similarities or differences.

# # 1 - high , 0 - low

Total with Low CRT - 572
           Died in 5 years - 77
           Proportion who died in 5 years - 77/572 = 0.1346154
           
           
           -----------------------
                  |  1    |  0
           -----------------------
  Low CRT         |  77   |  495
  
          -----------------------
          
          Odds - 77/495  - 0.1555556

## (e)
From the logistic regression model, what is the estimated odds of dying within 5 years for subjects with high creatinine levels.   What is the estimated probability of dying within 5 years from the logistic regression model for subjects with high creatinine levels ? 

Odds = 0.1562 * 2.367 = 0.3697254
- Using  prob= odds / (1+odds): 0.3697254/(1+0.3697254) = 0.2699267


#### Can answer the question probability of not dying in 5 given high creatnine levels

## (f)
For the subset of subjects in the sample with high creatinine levels, calculate the proportion who died within 5 years. Also calculate the sample odds of dying within 5 years for subjects with high creatinine levels.  Compare the sample proportions (or probabilities) and sample odds  to the corresponding estimates in part 1e from the logistic regression model? Briefly explain any similarities or differences.

# # 1 - high , 0 - low

Total with high CRT - 165
           Died in 5 years - 44
           Proportion who died in 5 years - 44/165 = 0.2666667
           
           
           -----------------------
                  |  1    |  0
           -----------------------
  High CRT        |  44   |  121
  
          -----------------------
          
          Odds - 1- (44+121)/121 = 1- 1.36 = .36
          (OR)
          44/121  - 0.3636364


## (g)
Give full inference regarding an association  between 5 year all-cause mortality and serum creatinine levels from the logistic regression model with  indicator of death within 5 years as the response and indicator of high creatinine level as the predictor.  

#### explanation of slope -  odds ratio


## (h)
How would your answers to part b change if you were instead asked to fit a logistic regression model with indicator of death within 5 years as the response variable, but with indicator of low serum creatinine level as the predictor?  Would the statistical evidence for an association between 5 year mortality and serum creatinine levels change?  Briefly explain.


## (i)
How would your answers to part b change if you were instead asked fit a logistic regression model with indicator of surviving  at least 5 years as the response variable  and  indicator of high creatinine level as the predictor?  Would the statistical evidence for an association between 5 year mortality and serum creatinine levels change?  Briefly explain.   



# Question 2
In question 1, a prospective association analysis was conducted where we investigated differences in the distribution of death within 5 years across groups defined by serum creatinine level.  In this question, you will now conduct a retrospective analysis and fit a logistic regression model for the distribution of serum creatinine across groups defined by vital status at 5 years.  In your retrospective logistic regression model,  use an indicator for high serum creatinine level as the response, and indicator of death within 5 years as the predictor.  (Only provide a formal report of inference when asked to.)


```{r,echo=FALSE,include=FALSE}
### Question 1:
# Part (a)
logiscrtl2 <- glm(crtind~deathin5,family = binomial,data = mri)
summary(logiscrtl2)

#Call:
#glm(formula = crtind ~ deathin5, family = binomial, data = mri)

#Deviance Residuals: 
#    Min       1Q   Median       3Q      Max  
#-0.9508  -0.6576  -0.6576  -0.6576   1.8098  

#Coefficients:
#             Estimate Std. Error z value Pr(>|z|)    
#(Intercept)   -1.4214     0.1021 -13.917  < 2e-16 ***
#deathin5TRUE   0.8618     0.2148   4.012 6.03e-05 ***
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#(Dispersion parameter for binomial family taken to be 1)

#    Null deviance: 776.83  on 732  degrees of freedom
#Residual deviance: 761.57  on 731  degrees of freedom
#  (2 observations deleted due to missingness)
#AIC: 765.57

#Number of Fisher Scoring iterations: 4

#> exp(-1.4214)
#[1] 0.2413759
#> exp(0.8618)
#[1] 2.367418
 
```

## (a) 
Provide an interpretation of the slope and the intercept in your regression model in terms of the response variable (indicator of high creatinine level) and the predictor variable (indicator of death within 5 years).  

Our model here is  log odds = -1.4214 + 0.8618  * Deathin5i

when predictor is 0(Does not die in 5 years) –Found by exponentiation of the intercept from the logistic regression: exp(b0) = 0.2413759

Odds ratio between groups differing in the value of the predictor by 1 unit–Found by exponentiation of the slope from the logistic regression: exp(b1) = 2.367418


```{r}
### Question 1:

```

## (b)
From the logistic regression model,  what is the estimated odds of high creatinine level for subjects who die within 5 years?   What is the estimated probability of having high  serum creatinine for subjects who die within 5 years.

```{r}
### Question 1:

#exp(-1.4214 + (0.8618  * 1))
#[1] 0.5714376

# 0.5714376 / (1+0.5714376)
```


Odds is 0.57
Using  prob = odds / (1+odds): 0.36364

## (c)
From the logistic regression model, what is the estimated odds of having a high creatinine level for subjects who survive at least 5 years?   What is the estimated probability of having a high serum creatinine for subjects who survive at least 5 years.

```{r}

#exp(-1.4214 + (0.8618  * 0))
#[1] 0.2413759

# 0.2413759 / (1+0.2413759)
# [1] 0.1944422
```

Odds is 0.2413759
Using  prob= odds / (1+odds): 0.19

## (d)
Give full inference regarding an association  between 5 year all-cause mortality and serum creatinine levels from the logistic regression model with  indicator of high serum creatinine as the response and an indicator of death within 5 years as the predictor.  

## (e)
Compare the association results in part 2d from the retrospective logistic model to the association results in part 1g from the prospective logistic regression model.  Briefly describe any similarities or differences.  


# Question 3

Perform a regression analysis to evaluate  an association between odds of death within 5 years and   the continuous measure of serum creatinine levels (i.e., do not use a dichotomized variable for serum creatinine levels in this analysis).  
## (a)

Provide an interpretation of the slope and the intercept in your logistic regression model.

```{r ,echo=FALSE, include=FALSE}
### Question 2:
# note: round() Does not print trailing zeros
regress("odds",formula = deathin5 ~ crt, data = mri)

#Call:
#regress(fnctl = "odds", formula = deathin5 ~ crt, data = mri)

#Deviance Residuals: 
#    Min       1Q   Median       3Q      Max  
#-1.0917  -0.5964  -0.5492  -0.4643   2.2837  

#Coefficients:

#Raw Model:
#                 Estimate  Naive SE  Robust SE       F stat    df Pr(>F)   
#[1] Intercept     -3.605     0.4012    0.4008            80.89 1  < 0.00005
#[2] crt            1.789     0.3398    0.3325            28.97 1  < 0.00005

#Transformed Model:
#                 e(Est)    e(95%L)   e(95%H)         F stat    df Pr(>F)   
#[1] Intercept     0.02719   0.01238   0.05973            80.89 1  < 0.00005
#[2] crt            5.986     3.116     11.50             28.97 1  < 0.00005

#(Dispersion parameter for binomial family taken to be 1)

#    Null deviance: 656.75  on 732  degrees of freedom
#Residual deviance: 624.70  on 731  degrees of freedom
#  (2 observations deleted due to missingness)
#AIC: 628.7

#Number of Fisher Scoring iterations: 4


#> exp(-3.605)
#[1] 0.02718744
#> exp(2.367418)
#[1] 10.66981

```
The model we get is log odds of DeathIn5 = -3.605 + 1.789 * CRT

Intercept, when predictor is 0(Creatinine level) – Found by exponentiation of the intercept from the logistic regression: exp(b0) =  0.02718744

Odds ratio between groups differing in the value of the predictor by 1 unit–Found by exponentiation of the slope from the logistic regression: exp(b1) = 10.66981

* Two rows deleted because of NA

## (b)
Give full inference for an association  between 5 year all-cause mortality and  serum creatinine levels from your logistic regression model. 

```{r,include=FALSE}
### Question 2:
# note: round() Does not print trailing zeros
#(e-.00006242 , e0.007396 ) = (1.051,1.056)
```

eg
From logistic regression analysis, we estimate that for two groups that differ by one unit(mg/dl) in CRT, the odds of death in 5 years is  10.6% higher in the group that differs by 1 unit(mg/dl), also, this estimate is found to be statistically significant. A 95% CI suggests that this observation is not unusual if a group that is one year older might have odds of stroke that was anywhere from 3.1% lower or 11.5% higher than the group that has a higher CRT level . A two-sided p value of 0.00005 suggests that we can with high confidence reject the null hypothesis that the odds of death in 5 years is not associated with CRT. 



