---
title: "Homework 4"
author: "Aji John"
date: "Winter Quarter 2018"
output:
  pdf_document: default
  word_document: default
---



```{r setup, include=FALSE}
### Setting up the packages, options we'll need:
library(knitr)
library(ggplot2)
library(tidyverse)
library(uwIntroStats)
knitr::opts_chunk$set(echo = FALSE)

```


# Question 1
Perform a regression analysis for evaluating an association between serum creatinine level and 5 year all-cause mortality by comparing the odds of death within 5 years across groups defined by whether the subjects have “high”  or “low” creatinine levels, where serum creatinine levels greater than 1.2 are considered to be “high”, (i.e., “high” corresponds to  creatinine > 1.2 mg/dl and “low” corresponds to creatinine ≤ 1.2 mg/dl ).  In your logistic regression model, use an indicator of death within 5 years as the  response, and an indicator of high serum creatinine level as the predictor. (Only provide a formal report of inference when asked to.) 

```{r ,echo=FALSE,include=FALSE}
### Question 1:
# Part (a)
mri <- read.table("data/mri.txt",header=TRUE) 
mri$deathin5 <- ifelse(mri$obstime <= 365.25 * 5 & mri$death == 1, TRUE, FALSE) 
mri$death5 <- (mri$obstime < 5*365) & (mri$death == 1)

#create high/low for crt

mri$crtind <- ifelse(mri$crt>1.2,1,0)
mri$logcrt <- log(mri$crt)
#kable(descrip(mri[mri$death==0,'crt']), format = "markdown")
#kable(summary(cars), format = "markdown")
kable(mri%>%filter(!is.na(crt) & deathin5==FALSE) %>% 
        group_by(deathin5) %>%
        summarise(n=n(),sd=sd(crt),mean=mean(crt))%>%
        select(c('n','mean','sd')) %>%
        as.data.frame() , format = "markdown")
 

```

## (a)

Is this a saturated regression model? Explain your answer.

Yes, it is a saturated model, as number of parameters equal to number of coefficients.

## (b)

Provide an interpretation of the slope and the intercept in your regression model  in terms of the response variable (indicator of death within 5 years) and the predictor variable (high  creatinine).

Odds when predictor is 0–Found by exponentiation of the intercept from the logistic regression: exp(b0)•Odds ratio between groups differing in the value of the predictor by 1 unit–Found by exponentiation of the slope from the logistic regression: exp(b1)

## (c)
From the logistic regression model, what is the estimated odds of dying within 5 years for subjects with low creatinine levels.   What is the estimated probability of dying within 5 years from the logistic regression model for subjects with low creatinine levels ? 

```{r,echo=FALSE,include=FALSE}
### Question 1:
# Part (a)
logiscrtl <- glm(deathin5 ~ crtind,family = binomial,data = mri)
summary(logiscrtl)

#Call:
#glm(formula = deathin5 ~ crtind, family = binomial, data = mri)

#Deviance Residuals: 
#    Min       1Q   Median       3Q      Max  
#-0.7933  -0.5387  -0.5387  -0.5387   2.0009  

#Coefficients:
#            Estimate Std. Error z value Pr(>|z|)    
#(Intercept)  -1.8567     0.1225 -15.152  < 2e-16 ***
#crtind        0.8618     0.2148   4.012 6.03e-05 ***
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#(Dispersion parameter for binomial family taken to be 1)

#    Null deviance: 656.75  on 732  degrees of freedom
#Residual deviance: 641.50  on 731  degrees of freedom
#  (2 observations deleted due to missingness)
#AIC: 645.5

#Number of Fisher Scoring iterations: 4

# exp(--1.8567 )
#[1] 6.402573
#> 6.402573/(1+6.402573)
#[1] 0.8649118
 
```

Odds of dying within 5 years for subjects with low creatinine levels is e-1.8567=  6.402573

Probability of dying within 5 years for subjects with low creatinine levels is 0.8649118.

 - Using  prob= odds / (1+odds):  6.402573 / (1+6.402573)= 0.8649118.

## (d)
For the subset of subjects in the sample with low serum creatinine, calculate the proportion who died within 5 years. Also calculate the sample odds of dying within 5 years for subjects with low creatinine levels.  Compare the sample proportion and sample odds  to the corresponding estimates in part 1c from the logistic regression model? Briefly explain any similarities or differences.

Total with Low CRT - 165
           Died in 5 years - 44
           Proportion who died in 5 years - 44/165 = 0.2666667
           
           
           -----------------------
                  |  0    |  1
           -----------------------
  Low CRT         |  44   |  121
  
          -----------------------
          
          Odds - 1- (44+121)/121 = 1- 1.36 = .36

## (e)
From the logistic regression model, what is the estimated odds of dying within 5 years for subjects with high creatinine levels.   What is the estimated probability of dying within 5 years from the logistic regression model for subjects with high creatinine levels ? 

## (f)
For the subset of subjects in the sample with high creatinine levels, calculate the proportion who died within 5 years. Also calculate the sample odds of dying within 5 years for subjects with high creatinine levels.  Compare the sample proportions (or probabilities) and sample odds  to the corresponding estimates in part 1e from the logistic regression model? Briefly explain any similarities or differences.

## (g)
Give full inference regarding an association  between 5 year all-cause mortality and serum creatinine levels from the logistic regression model with  indicator of death within 5 years as the response and indicator of high creatinine level as the predictor.  

## (h)
How would your answers to part b change if you were instead asked to fit a logistic regression model with indicator of death within 5 years as the response variable, but with indicator of low serum creatinine level as the predictor?  Would the statistical evidence for an association between 5 year mortality and serum creatinine levels change?  Briefly explain.

## (i)
How would your answers to part b change if you were instead asked fit a logistic regression model with indicator of surviving  at least 5 years as the response variable  and  indicator of high creatinine level as the predictor?  Would the statistical evidence for an association between 5 year mortality and serum creatinine levels change?  Briefly explain.     

# Question 2
In question 1, a prospective association analysis was conducted where we investigated differences in the distribution of death within 5 years across groups defined by serum creatinine level.  In this question, you will now conduct a retrospective analysis and fit a logistic regression model for the distribution of serum creatinine across groups defined by vital status at 5 years.  In your retrospective logistic regression model,  use an indicator for high serum creatinine level as the response, and indicator of death within 5 years as the predictor.  (Only provide a formal report of inference when asked to.)
 
## (a) 
Provide an interpretation of the slope and the intercept in your regression model in terms of the response variable (indicator of high creatinine level) and the predictor variable (indicator of death within 5 years).  

```{r}
### Question 1:
# Part (a)
plot(data=mri,crt~age)
abline(lm(crt~age, data = mri),col="red")

```

## (b)
From the logistic regression model,  what is the estimated odds of high creatinine level for subjects who die within 5 years?   What is the estimated probability of having high  serum creatinine for subjects who die within 5 years.

## (c)
From the logistic regression model, what is the estimated odds of having a high creatinine level for subjects who survive at least 5 years?   What is the estimated probability of having a high serum creatinine for subjects who survive at least 5 years.

## (d)
Give full inference regarding an association  between 5 year all-cause mortality and serum creatinine levels from the logistic regression model with  indicator of high serum creatinine as the response and an indicator of death within 5 years as the predictor.  

## (e)
Compare the association results in part 2d from the retrospective logistic model to the association results in part 1g from the prospective logistic regression model.  Briefly describe any similarities or differences.  


# Question 3

Perform a regression analysis to evaluate  an association between odds of death within 5 years and   the continuous measure of serum creatinine levels (i.e., do not use a dichotomized variable for serum creatinine levels in this analysis).  
## (a)

Provide an interpretation of the slope and the intercept in your logistic regression model.

```{r ,include=FALSE}
### Question 2:
# note: round() Does not print trailing zeros
regress("mean",formula = logcrt ~ age, data = mri)


```


## (b)
Give full inference for an association  between 5 year all-cause mortality and  serum creatinine levels from your logistic regression model. 

```{r,include=FALSE}
### Question 2:
# note: round() Does not print trailing zeros
#(e-.00006242 , e0.007396 ) = (1.051,1.056)
```

eg
From logistic regression analysis, we estimate that for two groups that differ by one year in age, the odds of stroke is  3.4% higher in the older group, though this estimate is not statistically significant. A 95% CI suggests that this observation is not unusual if a group that is one year older might have odds of stroke that was anywhere from 0.8% lower or 7.8% higher than the younger group.” A two-sided p value of 0.113 suggests that we can not with high confidence reject the null hypothesis that the odds of stroke are not associated with age. 



